---
title: "Top-level functions"
author: "Cas Neyens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{top_level}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r setup, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(RMODFLOW)
```

RMODFLOW provides `rmf_create_*`, `rmf_read_*` and `rmf_write_*` functions for individual MODFLOW packages. In most cases however, it is useful to read, write and create entire MODFLOW models. For this, RMODFLOW provides the top-level `rmf_create`, `rmf_read` and `rmf_write` functions. The main advantage of using these top-level functions, besides the obvious less verbose code, is that all options are set correctly: free/fixed formats, binary, additional arguments etc. are all set correctly by these functions. 

## Read
To read in an entire model, use `rmf_read()` on the MODFLOW name file. This will read in all the input packages active during the simulation. The only additional argument that has to be specified by the user is the `precision` of binary files ("single" by default).
```{r}
rma <- rmf_example_file("rocky-mountain-arsenal.nam") %>%
  rmf_read()
```

To also include output files, set `output = TRUE` but be warned: this might become slow if output files are large.
To suppress the writing of information to the console, set `verbose = FALSE`.

`rmf_read` returns a list of class `modflow` which holds all the packages. It can be useful to "unpack" this list with `list2env(model, .Globalenv)`.

## Create
To create a RMODFLOW `modflow` object, use `rmf_create()`. This function takes as input RMODFLOW objects of class `rmf_package` or a single list with `rmf_package` objects. If a `nam` object is not provided, it is created (silently). To overwrite the cell-by-cell flow package number of all objects that have a this number, specify a `cbc` value.
```{r}
modflow <- rmf_create(rma$dis, rma$bas, rma$lpf, rma$oc, rma$pcg, rma$wel, rma$chd, cbc = 88)
str(modflow)
```

All MODFLOW simulations must have at least a DIS, BAS, solver and flow package. If any of those are missing or duplicated, `rmf_create` will throw an error.
```{r, error=T, purl=FALSE}
modflow <- rmf_create(rma$dis, rma$bas, rma$lpf, rma$oc)
```

## Write
To write a model, use `rmf_write()` and pass a `modflow` object and the path to a MODFLOW name file. 
```{r, eval = FALSE}
rmf_write(modflow, file = "input.nam", verbose = FALSE)
```

You can exclude certain packages from writing by supplying an `exclude` argument with their RMODFLOW abbreviations. This can be useful if you e.g. want try out excluding a package from the simulation without recreating the entire `modflow` object. Because this bypasses the creation of the `modflow` object, you can exclude required packages such as the solver or the DIS file without RMODFLOW erroring out, so be careful.
```{r, eval = FALSE}
rmf_write(modflow, file = "input.nam", verbose = FALSE, exclude = c("wel", "chd"))
```

